On randomness and gravel rivers: Einstein and Bagnold in modern context

### 2021-09-03 revision/ch6conc.tex
added phd thesis.

The solutions of the master equation in a given problem produce the probability distribution for the response variable of interest, which in the thesis has included at different points the bedload particle position (chs. \ref{ch:flux} and \ref{ch:downDiff}), the particle velocity (ch. \ref{ch:langevin}), the bed elevation (ch. \ref{ch:ch3}), and the sediment transport rate (ch. \ref{ch:flux}).
The same stochastic strategy should be applicable to a wide range of problems in geomorphology where phenomena can built up from component parts with apparently noisy characteristics.

@@

A major challenge in this research is that only a handful of noises in statistical physics are comprehensively understood \citep{Horsthemke1984}, so the available options in stochastic modelling are constrained.
The noises used in this thesis include white Gaussian, Poisson, and dichotomous noises, representing erratic fluctuations, sequences of spikes, and random switches respectively \citep{VanDenBroeck1983}.
Turbulence was described using Gaussian noise, while instantaneous steps, particle arrivals to a control surface, and particle-bed collisions were described with Poisson noise. Alternation between motion and rest was represented with dichotomous noise.
Whenever these processes acted in combination, the dynamical equations describing the response variable included multiple sources of idealized noise as required.
This inclusion of multiple noise sources has not yet to my knowledge been pursued in any earlier stochastic models of sediment transport.

River science offers no guarantee that these idealized white noises which we happen to understand best are sufficient to describe its phenomena.
White noises are an idealization which is probably never realized in nature \citep{Gardiner1983,Kubo1978}.
In contrast, colored noises in which fluctuations have favoured frequencies are well-known to occur in many fluid and granular physics phenomena, most famously in context of fluid turbulence \citep{Kolmogorov1941,Nikora2000} and granular collapse \citep{Bak1987,Jensen1998}.
In river science, many phenomena exhibit colored spectra, like the size distributions of bedforms \citep{Nikora1997,Guala2014}, the fluid forces on bed particles \citep{Dwivedi2011, Amir2014}, the roughness characteristics of gravel beds \citep{Aberle2006,Singh2012}, and sediment flux timeseries \citep{Dhont2018,Chartrand2021}.
It will be problematic if these phenomena require colored noise for their description. Dynamical equations driven by colored noise can be extremely challenging to solve \citep{Hanggi1978,Luczka2005,Hanggi2007}.
Even if white noise models like those developed in this thesis are not exactly accurate, they remain necessary as a basis for comparison when formulating and solving colored noise models \citep{Fox1986,Moss1989}.

@@

\citet{Ballio2018} explained that scale-dependence originates from individual particle trajectories, but this had not been described in a mathematical model until now.
Descriptions of the mean sediment flux from the movement characteristics of individual grains have existed now for a long time (sec. \ref{sec:einflux}), and an emerging body of research has producing the full probability distribution of the flux, but without referencing individual movement characteristics (secs. \ref{sec:birthdeath}-\ref{sec:renewal}). 
This work unifies these two research themes and presents a statistical mechanics formulation of bed load sediment transport based on individual particle trajectories.

@@

Fourth, chapter \ref{ch:downDiff} presenting a model of sediment trajectories through motion, rest, and burial, describing sediment transport across local, intermediate, global, and geomorphic ranges (), and producing new understanding of how exactly the distinct spreading characteristics of particles within each of these ranges arise \citep[e.g][]{Pretzlav2021}.

@@

\subsection{Landscape dynamics and channel morphology}

The first limitation concerns the assumption, implicit in every chapter of this thesis, that sediment transport characteristics are steady and uniform in time and space.
This assumption contrasts with conditions in real gravel-bed rivers, where riffles, bars, and steps coordinate the movements of individual grains \citep{Ashmore1998,McDowell2020}, sediment is supplied in episodic bursts from mass movements \citep{Benda1990, Muller2018}, woody debris coordinates sediment movement \citep{Eaton2012,Reid2019}, and variable flow conditions modify bed texture and sediment availability \citep{Mao2012,Phillips2018}.

@@

The second major limitation of this work concerns grain size.
In actuality, sediments in rivers spans a range of sizes, and differential mobility based on particle size produces spatial sorting, both vertically as in bed armoring \citep{Parker1982,Wilcock1989,Aberle2006}, and laterally as patch, particle cluster, or riffle development \citep{Nelson2014,Venditti2017}.

@@

Earth's landforms are shaped by numerous transport phenomena from booming debris flows \citep{Iverson1997} to barely perceptible hillslope creep \citep{Deshpande2021}.
Different phenomena are basically distinguished by the relative importance of fluid, granular, and gravitational forces in sustaining them \citep{Jerolmack2019}.

Weak bedload transport is characterized by fluid forces small in comparison to gravity, and collision forces comparable to gravity.
Viewed in this way, the work in this thesis targets a minute region of the vast parameter space spanned by Earth's geophysical flows, and geomorphology requires characterization of them all.
Future studies should continue the effort \citep{Furbish2021a} to describe these processes as different expressions of the same basic statistical mechanics building blocks.

@@

This thesis described the movements of individual grains along streambeds using probabilistic methods.
The research has related the overall sediment transport rates responsible for channel evolution to the movements of individual grains.
At base level, the thesis embraces variability as an intrinsic part of Earth surface dynamics, and it produces descriptions which predict mean values as well as the magnitude of their fluctuations.

The founders of process geomorphology always acknowledged the role of variability in landscape evolution \citep{Horton1945,Strahler1952,Langbein1964}, although their main efforts were to develop strategies to describe landscapes without including it, like averaged shear stresses to avoid fluid turbulence \citep{MeyerPeter1948,Bagnold1954}, competent conditions to replace climate fluctuations \citep{Wolman1959,Wolman1978}, and representative grain sizes to avoid evolving grain size distributions \citep{Parker1982,Andrews1983}.

>>>>

The solutions of the master equation in a given problem produce the probability distribution for the response variable of interest, which in the thesis has included at different points the bedload particle position (chs. \ref{ch:flux} and \ref{ch:downDiff}), the particle velocity (ch. \ref{ch:langevin}), the bed elevation (ch. \ref{ch:ch3}), and the sediment transport rate (ch. \ref{ch:flux}).
The same stochastic strategy should be applicable to a wide range of problems in geomorphology where phenomena can be built up from component parts with apparently noisy characteristics.

@@

A major challenge in this research is that only a handful of noises in statistical physics are comprehensively understood \citep{Horsthemke1984}, so the available options in stochastic modelling are constrained.
The noises used in this thesis include white Gaussian, Poisson (shot), and dichotomous noises, representing erratic fluctuations, sequences of spikes, and random switches respectively \citep{VanDenBroeck1983}.
Turbulence was described using Gaussian noise, while instantaneous steps, particle arrivals to a control surface, and particle-bed collisions were described with Poisson noise. Alternation between motion and rest was represented with dichotomous noise.
Whenever these processes acted in combination, the dynamical equations describing the response variable included multiple sources of idealized noise as required.
This inclusion of multiple noise sources has not yet to my knowledge been pursued in any earlier stochastic models of sediment transport.

River science offers no guarantee that these idealized white noises which we happen to understand best are sufficient to describe its phenomena.
White noises are an idealization which is probably never realized in nature \citep{Gardiner1983,Kubo1978}.
In contrast, colored noises in which fluctuations have favoured frequencies are well-known to occur in many fluid and granular physics phenomena, most famously in fluid turbulence \citep{Kolmogorov1941,Nikora2000} and granular collapse \citep{Bak1987,Jensen1998}.
In river science, many phenomena exhibit colored spectra, like the size distributions of bedforms \citep{Nikora1997,Guala2014}, the fluid forces on bed particles \citep{Dwivedi2011, Amir2014}, the roughness characteristics of gravel beds \citep{Aberle2006,Singh2012}, and sediment flux timeseries \citep{Dhont2018,Chartrand2021}.
It will be problematic if these phenomena require colored noise for their description. Dynamical equations driven by colored noise can be extremely challenging to solve \citep{Hanggi1978,Luczka2005,Hanggi2007}.
Even if white noise models like those developed in this thesis are not exactly accurate, they remain necessary as a basis for comparison when formulating and solving colored noise models \citep{Fox1986,Moss1989}.

@@

\citet{Ballio2018} explained that scale-dependence originates from individual particle trajectories, but this had not been described in a mathematical model until now.
Descriptions of the mean sediment flux from the movement characteristics of individual grains have existed now for a long time (sec. \ref{sec:einflux}), and an emerging body of research has produced the full probability distribution of the flux without referencing individual movement characteristics (secs. \ref{sec:birthdeath}-\ref{sec:renewal}). 
This work unifies these two research themes and presents a statistical mechanics formulation of bed load sediment transport based on individual particle trajectories.

@@

Fourth, chapter \ref{ch:downDiff} presents a model of sediment trajectories through motion, rest, and burial, describing sediment transport across local, intermediate, global, and geomorphic ranges (sec. \ref{sec:trajmodels}), and producing new understanding of how exactly the distinct spreading characteristics of particles within each of these ranges arise \citep[e.g][]{Pretzlav2021}.

@@

\subsection{Channel dynamics and morphology}

The first limitation concerns the assumption, implicit in every chapter of this thesis, that sediment transport characteristics are steady and uniform in time and space.
This assumption contrasts with conditions in real gravel-bed rivers, where riffles, bars, and steps coordinate the movements of individual grains \citep{Ashmore1998,McDowell2020}, sediment is supplied in episodic bursts from mass movements \citep{Benda1990, Muller2018}, woody debris directs sediment accumulation \citep{Eaton2012,Reid2019}, and variable flow conditions modify bed texture and sediment availability \citep{Mao2012,Phillips2018}.

@@

The second major limitation of this work concerns grain size.
In actuality, sediments in rivers span a range of sizes, and differential mobility based on particle size produces spatial sorting, both vertically as in bed armoring \citep{Parker1982,Wilcock1989,Aberle2006}, and laterally as patch, particle cluster, or riffle development \citep{Nelson2014,Venditti2017}.

@@

Earth's landforms are shaped by numerous transport phenomena from booming debris flows \citep{Iverson1997} to barely perceptible hillslope creep \citep{Deshpande2021}.
Different transport phenomena are basically distinguished by the relative importance of fluid, granular, and gravitational forces in sustaining them \citep{Jerolmack2019}.

Weak bedload transport is characterized by fluid forces small in comparison to gravity, and collision forces comparable to gravity.
Viewed in this way, the work in this thesis targets a minute region of the vast parameter space spanned by Earth's geophysical flows, and geomorphology requires characterization of them all.
Future studies should continue the effort \citep[e.g.][]{Furbish2021a} to describe these processes as different expressions of the same basic statistical mechanics building blocks.

@@

This thesis has described the movements of individual grains along streambeds using probabilistic methods.
The research relates the overall sediment transport rate to the movements of individual grains.
At base level, the thesis embraces variability as an intrinsic part of Earth surface dynamics, and it produces descriptions which predict mean values as well as the magnitude of their fluctuations.

The founders of process geomorphology always acknowledged the role of variability in landscape evolution \citep{Horton1945,Strahler1952,Langbein1964}, although their main efforts were to develop strategies to describe landscales without including variability, like averaged shear stresses to avoid fluid turbulence \citep{MeyerPeter1948,Bagnold1954}, competent conditions to replace climatic fluctuations \citep{Wolman1959,Wolman1978}, and representative grain sizes to avoid evolving surface grain size distributions \citep{Parker1982,Andrews1983}.

### 2021-09-03 revision/formerPapers/com-piece/scraps/com-piece0.tex
added phd thesis.

\title{On argumentation and gravel rivers: Einstein and Bagnold in a modern context}
\author{Kevin Pierce}
\maketitle

%% intro

Gravel bed rivers are troubled natural systems, and changes in their morphology are linked to a wide array of ecological and engineering problems.
Since the transport of coarse material as bedload is an important control on channel morphology, bedload transport has been the subject of intense research for over a century. 
Despite this effort, our ability to predict bedload transport rates is inadequate.  
In natural streams, predictions of bedload transport can deviate from measured values by factors of 100 or 1000. 
Since the bedload transport process is governed by Newtonian mechanics, it should be fully characterized by exchanges of momentum between the granular and fluid phases, highlighting a mechanical description as an attractive prospect.
However, because bedload transport results from a turbulent flow interacting with granular matter, the full mechanical description is a fusion of two notoriously difficult and tenuously understood subfields of classical physics, meaning many of the required details are missing.
In absence of these details, many researchers have attempted to reverse engineer mechanical theories of bedload transport, and among these, two of them have been taken up by generations of researchers, forming two main research paradigms in the physics of bedload transport. 
These are the stochastic approach, initiated by \citet{Einstein1937, Einstein1950, Einstein1964} and the deterministic approach initiated by \citet{Bagnold1956, Bagnold1966, Bagnold1973}.

Each of these approaches is founded on the application of mechanical reasoning to a simplified concept of bedload motion, characterized by empirical parameters, so that both categories of models must be calibrated to applications. 
Also, the simplified concepts of Einstein and Bagnold-type models are different with regard to most aspects of bedload motion.
Accordingly, when proponents of different schools discuss sediment transport processes, they often find miscommunications and disagreements. 
This conflict is exascerbated by observations that neither mechanical approach is, practically speaking, very good; and that neither approach is clearly superior than the other. 
Specifically, both of these mechanical paradigms are expected to generate poorer bedload transport predictions than simpler power law curves, calibrated to empirical data without reference to mechanical concepts \citep{Barry2004}; and the semi-empirical aspect of these models provides flexibility, so comparing them by predictive ability is difficult, as neither approach shows uniform superiority across a full range of flow and sediment characteristics. 
Accordingly, although each mechanical approach has clear merits and demerits, and it's safe to conclude neither is perfect; the perspective that either is fully superior, for any reason apart from practicality, lacks objective support. 
Because there is no clear path to join the two schools together, proponents of each school toil away in their respective camps, solving problems within their own simplified conceptual pictures, and failing to communicate science across the stochastic / deterministic rift that Einstein and Bagnold have left behind. 
We need to ask: is this rift justified by evidence? Are Bagnold and Einstein really so different that they share no common ground, and is one or the other completely wrong? If so, theorists should out the failed approach and concentrate on its successor together. If not, we should work together to synthesize something new out of the strongest elements of each paradigm.

To answer these questions, we should carefully assess the concepts of each paradigm, and try to discern any conditions of bedload transport under which one description provides more or less useful information than the other. 
Because bedload transport rates are the cumulative result of individual particle motions, an appropriate starting point seems to be a clear physical picture of individual bedload motions, and this picture should be consistent with contemporary experiments. 
Bedload grains typically move intermittently, rolling, sliding, and bouncing along the river bed \citep{Einstein1950, Bagnold1973}.
% resting, entrainment by impulse, motion in turbulent flow, collisions with bed, deposition

% explain how bedload transport has different behaviors at low and high stages of bedload transport

% introduce the einstein theory, its revisions, and experiments which support it

% introduce the bagnold theory, its revisions, and experiments which support it

% highlight that bagnold is the high transport stage limit of tau^(3/2), while einstein is really the low transport stage limit where average properties of the flow and grains do not characterize bedload transport

% really, we should we should be joining these theories to describe wolman's intermediate region, which is where channel morphology happens

% however, in intermediate regions, the linkage between morphology and transport becomes more highly coupled; so we really need to integrate knowledge from higher scales with the stochastic theories of einstein and the power dissipation arguments of bagnold -- neither is good enough

% conclude by wrapping problems, stochastic, determinitsic, a synthetic description of intermediate conditions, where feedbacks are accentuated with morphology, with an interesting conclusion about scientific divides and tribalism impeding progress

Sediment grains within gravel-bed rivers typically move intermittently, rolling, sliding, and bouncing along the river bed, in a condition called bedload motion \citep{Einstein1950, Bagnold1973}.
The bedload motion of an individual grain is an alternating series of movements and rests, with these dynamics undoubtly governed by a Newtonian force balance between the solid and fluid phases \citep{Schmeeckle2013, Ji2013}.
However, the forces making up the balance are not exactly known in any context, and the granular \citep{Lamb2008, Frey2011, Booth2014} and fluid \citep{Schmeeckle2007, Celik2014, Amir2014, Shih2017} forces are known to exhibit considerable variability in magnitude, direction, and duration. 
Looking at experiments on the mechanics of bedload transport, it's clear that the Einstein and Bagnold schools are simplified from reality.

We should really be attempting a synthesis between them, but to achieve this, we have to assess the experimental findings on the processes underlying bedload transport, as well as the concepts and development of the Einstein and Bagnold schools.  
To make this synthesis, we should be greedy, taking useful ideas from wherever we can find them. 
To this end, we can start from a question: which concepts from each school fit with experimental findings, and how can they be fused together for a predictive theory of bedload transport?

%% bedload experiments

% first describe broad brush ideas. initiation, motion, cessation 
To approach the question, we should develop a clear mechanical understanding of individual bedload motions which is supported by experiments, and we should consider where Einstein and Bagnold's theories fall in relation to it.  
Because bedload motion is intermittent, a common and useful concept is to divide individual particle motions into three stages: initiation, movement, and cessation. 
To interpret bedload experiments, investigators have studied how the properties of each stage result from a balance of granular and fluid interactions in accord with Newtonian mechanics. 
Initiation, or the transition of a bed particle from rest to motion, is undoubtedly a result of momentum transfer to the stationary particle, by the fluid and possibly by impacts from other moving grains, and these driving impulses are resisted by frictional and contact forces imparted on stationary grains by every other grain in contact. 
Similarly, the trajectories of particles, once they are in movement, are determined by the momentum transferred from the fluid phase and by impulses imparted by collisions with the irregular bed surface formed by stationary grains \citep{Wiberg1985, Bialik2015}. 
The specifics of the fluid and granular forces are not accessible: to fully characterize the fluid forces would require the full time series of pressure and velocity fields within the turbulent fluid, including within the porous bed, and to characterize the static forces and dynamic impulses imparted by the bed surface would require knowledge of the location and relative orientation of every grain within the bed. 
Accordingly, the initiation and trajectories of particles is very difficult to understand, but the cessasion of motion is even more difficult. 
Cessasion of motion, or the transition of a moving particle to rest, results from a chance coordination of upstream impact forces from the configuration of static grains with a lull in turbulent driving forces \citep{Pahtz2018}, again in accord with Newton's laws.

% now describe gaps in this understanding
This broad mechanical perspective on the processes underlying individual particle motions reveals many gaps in experimental knowledge. 
For one thing, in discussing the

% now describe the einstein theory, and

% initiation
% motion
% cessasion
% rates

%% Einstein and Bagnold

Motion of an individual particle is initiated when a sufficiently strong driving force is applied by the turbulent fluid over a timescale large enough to overcome the resistive forces imparted on that grain by frictional and normal responses from the every bed grain it contacts \citep{Paintal1971, Valyrakis2010, Celik2014}. 
In general, both the duration and magnitude of the applied fluid forces is important \citep{Diplas2008, Valkyrus2010, Celik2014}. 
This is an impulse concept, although in a viscous flow without velocity or pressure fluctuations, this impulse concept reduces to the classical perspective of a critical motion condition based upon velocity or shear stress \citep{Shields1936, Buffington1997}.

This initiation of motion by a turbulent impulse is undoubtedly governed by Newton's laws, although the forces making up the balance are not exactly known in any context, and the granular resistive \citep{Lamb2008} and fluid driving \citep{Schmeeckle2007, Celik2014, Amir2014, Shih2017} forces are known to exhibit considerable variability in magnitude, direction, and duration. 
Similarly, the trajectories of particles, once they are in motion, are determined by Newton's laws, contingent on the specifics of turbulent driving forces, including the full time series of turbulent forces, and by impulses imparted by collisions with the irregular bed surface formed by stationary grains \citep{Wiberg1985, Bialik2015}. 
The cessastion of motion is poorly understood, but presumably results from a chance coordination of local bed geometry with a lull in turbulent driving forces \citep{Pahtz2018}, again in accord with Newton's laws.  
Owing to these factors, a Newtonian description of bedload motion based on equations characterizing momentum exchange, is a difficult and probably impossible prospect. 
This difficulty results from the diverse phenomenology of the fluid and granular phases, and especially from the processes that hold them in feedback to induce bedload motion.

>>>>

\title{On randomness and gravel rivers: Einstein and Bagnold in modern context}
\author{Kevin Pierce}
\maketitle

% gravel bed rivers are in trouble 
% need to understand interaction of transport, storage, supply
% but transport has been tough to describe due to turbulence and granular interactions
% fundamentally the problem is to predict the rate given the water flow
% Currently, there are no general laws, and our ability to predict transport rates in natural streams is inadequate
% However, there are two prevalent research paradigms:

Gravel bed rivers are troubled natural features, and hosts to a huge set of ecological problems related to changes in their form and function.
These changes often link to human disturbances of the natural regimes of sediment transport, storage, and supply within these rivers. 
Accordingly, these processes are cornerstones of river science, and they have been carefully studied for well over a century. 
Despite this work, with regard to sediment transport in gravel bed rivers, there is very little practical progress. 
The earliest efforts were based upon correlating empirical data to power laws relating the rate of sediment transport to characteristics of the flow and sediment \citep{MeyerPeter1948}, and this work provided a foundation for the now classic approaches based on mechanistic physics. 
Among these early mechanistic approaches, two have been particularly influential, each initiating its own research school which are dominant today: these are the stochastic approach of \citet{Einstein1937, Einstein1950}, and the deterministic approach of \citet{Bagnold1956, Bagnold1973}.
A rift exists these two schools, and proponents of one face ire from the other, and they often find debate: these schools take different approaches to understand almost every aspect of sediment transport in gravel bed rivers. 
However, a reality underlies this divide: the most advanced mechanical theories of both schools, from a practical standpoint, remain less useful than modern theories based on fitting power laws to empirical data \citep{}, and apparently no theory among these provides reliable predictions of transport in natural streams \citep{}. 
Accordingly, we should take a step back and look at what's happened.

Among these, probably resulting from the heightened influence of fluid turbulence and granular interactions in gravel bed rivers, sediment transport has been a particularly difficult inquiry. 
Fundamentally, the problem is this: given the rate of water flow, the form of the channel, and the properties of the sediment grains within a gravel bed river, what is the rate of transport of sediment grains?
- describe how poor our capacity for prediction is 
This problem has been studied for over a century, with limited success, and no general laws. 
We need to understand why.

Sediment grains within gravel-bed rivers typically move intermittently, rolling, sliding, and bouncing along the river bed, a condition called bedload motion \citep{Einstein1950}.
The bedload motion of an individual grain is an alternating series of start to stop transitions, with these dynamics undoubtly governed by a Newtonian force balance between the solid and fluid phases.
However, the forces making up the balance are not exactly known in any context, and the granular \citep{Lamb2008} and fluid \citep{Schmeeckle2007, Celik2014, Amir2014, Shih2017} interactions are known to exhibit considerable variability in magnitude, direction, and duration. 
Owing to this complexity, and to our relative ignorance about the details of the underlying Newtonian basis of bedload transport, two main research schools have emerged, each leveraging a set of approximations and assumptions in order to make use of Newtonian concepts: these are the deterministic school, initiated by Bagnold, and based upon the concept of momentum exchange between the granular and fluid phases; and the stochastic school, initiated by Einstein, and based upon probabilistic transition rates between motion and rest states. 
In the river science community, these two schools are often opposed. 
In this essay, I argue this division is short-sighted, and probably finds more basis in feelings than facts.
Looking at the mechanical phenomena of bedload transport which experiments have revealed, it's clear that both stochastic and deterministic models of bedload transport are highly simplified away from the actual mechanics and processes at play.  
They both fail to predict bedload transport to similar degrees, although in different regions of the parameter space; and they each have important successes, of which we should take account. 
Accordingly, I argue we should really be synthesizing Bagnold and Einstein for a new understanding, and not counterposing them. 
This synthesis should happen in acknowledgement of the full complexity of the process, and with respect for experimental observations.

To begin with, we should discuss the experimentally observed phenomena underpinning bedload transport. 
Because bedload motion is intermittent, a useful division is to conceptualize it as three stages: initiation, movement, and cessation. 
The properties of each stage result from a balance of granular and fluid interactions. 
Motion of an individual particle is initiated when a sufficiently strong driving force is applied by the turbulent fluid over a timescale large enough to overcome the resistive forces imparted on that grain by frictional and normal responses from the every bed grain in contact \citep{Paintal1971, Valyrakis2010, Celik2014}. 
In general, the duration and magnitude of the applied fluid forces is important \citep{Diplas2008}; this is an impulse concept, although in a viscous flow without velocity or pressure fluctuations, this impulse concept reduces to the classical perspective of a critical motion condition based upon velocity or shear stress \citep{Shields1936, Buffington1997}. 
This initiation of motion by a turbulent impulse is undoubtedly governed by Newton's laws, although the forces making up the balance are not exactly known in any context, and the granular resistive \citep{Lamb2008} and fluid driving \citep{Schmeeckle2007, Celik2014, Amir2014, Shih2017} forces are known to exhibit considerable variability in magnitude, direction, and duration. 
Similarly, the trajectories of particles, once they are in motion, are determined by Newton's laws, contingent on the specifics of turbulent driving forces, including the full time series of turbulent forces, and by impulses imparted by collisions with the irregular bed surface formed by stationary grains \citep{Wiberg1985, Bialik2015}. 
The cessastion of motion is poorly understood, but presumably results from a chance coordination of local bed geometry with a lull in turbulent driving forces \citep{Pahtz2018}, again in accord with Newton's laws.  
Owing to these factors, a Newtonian description of bedload motion based on equations characterizing momentum exchange, is a difficult and probably impossible prospect. 
This difficulty results from the diverse phenomenology of the fluid and granular phases, and especially from the processes that hold them in feedback to induce bedload motion.
